---
title: "Bycatch and discards estimation for New Zealand fisheries (2018)"
author: "Charles T T Edwards (NIWA, Wellington, New Zealand)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: no
vignette: >
  %\VignetteIndexEntry{lhm}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path = 'fig/bde-', fig.width = 6, tidy = TRUE, tidy.opts = list(blank = TRUE, width.cutoff = 95), message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>")
```

```{r}
library(bde)
```

# Introduction

The `bde` package is designed to facilitate estimation of bycatch and discards in New Zealand fisheries using a two-part statistical model. The model is fitted to observer sampling data using the Bayesian `rstan` estimation framework. Estimated parameters are then used to predict the catch for unobserved commercial fishing effort.

The model uses a Biomial likelihood to fit to the aggregated count data $z_j$ by, for example, year ($j$):
\[
    z_j \sim B(n_j, \theta_j)
\]
whereas the positive catch data are included on a tow-by-tow basis. For observer record $i$, we therefore have:
\[
    y_i \sim LN(\mu_j, \sigma_j)
\]
which is necessary for estimation of the standard error term $\sigma_j$. The two-model parts are considered independent.

We illustrate utility of the package using simulated data.

